---
title: "Bridge to R-INLA and Stan"
date: "Created: 2023-12-13. Last modified: `r Sys.Date()`."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bridge to R-INLA and Stan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Using WCP prior in R-INLA

Our package allows user to provide WCP prior with a form that R-INLA software can use. Here we give an example with WCP$_2$ prior for $\phi$ of stationary AR(1) processes. The model setup and theoretical details can be found at Section 3.1 from https://arxiv.org/pdf/2312.04481.pdf. We will generate the WCP$_2$ prior by its analytic expression. Functions that encode analytic expression of WCP priors in our package have similar usage. For univariate prior, functions take a sequence of values of parameter of interests and for bivariate priors, functions take one pair of values of the two parameters. If users set the argument 'inla_table' to be TRUE, then the function will output the prior in the format that R-INLA can use.

We first simulate a stationary AR(1) process with length 10, standard deviation 0.1 and $\phi = 0.2$.
```{r}
# import INLA
library(WCPprior)
library(INLA)

# length of the AR1 process
n = 10
# standard deviation of the process
sigma = 0.1
# value of phi
phi = 0.2
# simulate the process
sim_data = numeric(n)
sim_data[1] = rnorm(1,mean = 0, sd = sigma)
for (i in 2:n){
 sim_data[i] = phi*sim_data[i-1] + rnorm(1, mean = 0, sd = sigma*sqrt(1-phi^2) )
}
data = data.frame(y = sim_data, t = 1:n)


```


The function 'WCP2_1D_AR1_analytic' in our package gives the true WCP$_2$ of $\phi \in [-1,1)$. 
```{r}
# create a sequence of values of phi
seq_phi = seq(from = -1, to = 0.99, by = 0.01)
# call the function and set inla_table = TRUE so that the function will return a prior that can be used by INLA
WCP2_prior_table = WCP2_1D_AR1_analytic (seq_phi = seq_phi, 
                               eta = 7,
                               n = 13,
                               sigma = sigma,
                               inla_table = TRUE)
# INLA formula
formula = y~ -1+f(t,model='ar1',hyper = WCP2_prior_table)
family = "Gaussian"
# fit the model by INLA
result = inla(formula=formula, data = data, family=family,
             control.predictor = list(compute=T))
summary(result)
# visualize the result
plot(data$y, col="blue",
     ylab="fitting result")
lines(result$summary.random$t[ ,"mean"])
```



Now we show a bivariate example with


