---
title: "Bridge to R-INLA and Stan"
date: "Created: 2023-12-13. Last modified: `r Sys.Date()`."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bridge to R-INLA and Stan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Using univariate WCP prior in R-INLA

Our package allows user to provide WCP prior with a form that R-INLA software can use. Here we give an example with WCP$_2$ prior for $\phi$ of stationary AR(1) processes. The model setup and theoretical details can be found at Section 3.1 from https://arxiv.org/pdf/2312.04481.pdf. We will generate the WCP$_2$ prior by its analytic expression. Functions that encode analytic expression of WCP priors in our package have similar usage. For univariate prior, functions take a sequence of values of parameter of interests and for bivariate priors, functions take one pair of values of the two parameters. If users set the argument 'inla_table' to be TRUE, then the function will output the prior in the format that R-INLA can use.


We now load the R-INLA and WCPprior library.
```{r, message=FALSE}
library(WCPprior)
library(INLA)


```



We first simulate a stationary AR(1) process with length 10. We set the standard deviation $\sigma = 0.1$ and $\phi = 0.2$.


```{r, message=FALSE}
# length of the AR1 process
n = 10
# standard deviation of the process
sigma = 0.1
# value of phi
phi = 0.2
# simulate the process
sim_data = numeric(n)
sim_data[1] = rnorm(1,mean = 0, sd = sigma)
for (i in 2:n){
 sim_data[i] = phi*sim_data[i-1] + rnorm(1, mean = 0, sd = sigma*sqrt(1-phi^2) )
}
data = data.frame(y = sim_data, t = 1:n)

```


Now we are ready to setup the WCP$_2$ prior of $\phi$ by using the function 'WCP2_1D_AR1_analytic' from WCPprior package. User should input a sequence of values of  $\phi \in [-1,1]$ with information of $n$, $\sigma$ and $\eta$. 
```{r}
# create a sequence of values of phi
seq_phi = seq(from = -1, to = 0.999, by = 0.001)
# call the function and set inla_table = TRUE so that the function will return a prior that can be used by INLA
WCP2_prior_table = WCP2_1D_AR1_analytic (seq_phi = seq_phi, 
                               eta = 15,
                               n = 10,
                               sigma = sigma,
                               inla_table = TRUE)



```
By setting 'inla_table = TRUE', the output of this function can be directly used by INLA for specifying prior in its formula.
```{r}
formula = y~ -1+f(t,model='ar1',hyper = WCP2_prior_table)
```
Then, we are ready to fit the model with the simulated process and WCP prior by INLA.
```{r}


result = inla(formula=formula, data = data, family="Gaussian",
             control.predictor = list(compute=T))

summary(result)
```

Finally, we visualize the result.

```{r}

plot(data$y, col="blue",
     ylab="fitting result")
lines(result$summary.random$t[ ,"mean"])
```


## Using univariate WCP prior in stan

WCPprior package also provides format of univariate WCP priors that can be used by stan. The implemented priors are: 
(1), WCP$_2$ prior for $\phi$ of stationary AR(1) process. (WCP2_1D_AR1_analytic) \
(2), WCP$_1$ prior for $\xi$, the tail index, of generalized Pareto distribution. (WCP1_1D_GPtail_analytic) \
(3), WCP$_2$ prior for precision parameter (reciprocal of variance) of Gaussian distribution. (WCP2_1D_Gaussian_precision_analytic) \
(4), WCP$_2$ prior for mean parameter of Gaussian distribution. (WCP2_1D_Gaussian_mean_analytic) \


Here we show an example of using 'WCP2_1D_AR1_analytic' function in stan.

The first block in the stan file should be a function block that includes the implemented functions from WCPprior package.
```{stan,  output.var='model', eval=FALSE}
functions {
#include /include/WCP_analytic.stan
}
```

The second block specifies data. That is an AR(1) process, y,  with length $N$.
```{stan,  output.var='model', eval=FALSE}
data {
  int<lower=0> N;
  vector[N] y;
}
```

The third block specifies parameter of interests, which is $\phi$ and its range is $[-1,1]$
```{stan,  output.var='model', eval=FALSE}
parameters {
  real<lower = -1, upper = 1> phi;
}
```

The last block specifies model setup including log-likelihood of the process and log-likehood of the WCP prior. Here, the first argument of the function 'WCP2_1D_AR1_analytic' is length of the process, the second one is value of $\eta$ and the last one is value of $\sigma$ (standard deviation of the process).
```{stan,  output.var='model', eval=FALSE}
model {
  for (n in 2:N)
    y[n] ~ normal(phi * y[n-1], 0.1);
  phi ~ WCP2_1D_AR1_analytic(N, 2.172101, 0.1);
}
```

Let us now create the `stan` file and build the model. It will be stored in the `model` object.

```{stan, output.var='model'}
functions {
#include ../inst/stan/include/WCP_analytic.stan
}
data {
  int<lower=0> N;
  vector[N] y;
}
parameters {
  real<lower = -1, upper = 1> phi;
}
model {
  for (n in 2:N){
    y[n] ~ normal(phi * y[n-1], 0.1);
  }
  phi ~ WCP2_1D_AR1_analytic(N, 2.172101, 0.1);
}
```

Let us sample an AR1 process, where `n` is the length of the process:

```{r}
n = 100
sigma = 0.1
phi = 0.8
sim_data = numeric(n)
sim_data[1] = rnorm(1,mean = 0, sd = sigma)
for (i in 2:n){
  sim_data[i] = phi*sim_data[i-1] + rnorm(1, mean = 0, sd = sigma*sqrt(1-phi^2) )
}
data <- list(N = 100L, 
             y = sim_data)
```

We will now fit the model in `stan`. Observe that we are calling the `sampling` function as we have the model as an `R` object.

```{r, warning = FALSE, message=FALSE}
library(rstan)
fit_WCP_AR <- sampling(model, data = data, iter = 1000) 
```

Let us plot the fitted model:

```{r}
plot(fit_WCP_AR)
```





## Using bivariate WCP prior in stan

The following two bivariate WCP priors that are also implemented in stan functions in WCPprior package:
(1), Bivariate WCP$_2$ prior for mean and standard deviation parameter of Gaussian distribution. (WCP2_2D_Gaussian_analytic) \
(2), Bivariate WCP$_1$ prior for $\sigma$ and $\xi$ of generalized Pareto distribution. (WCP1_2D_GP_analytic) \

Here we show an example of using 'WCP1_2D_GP_analytic' function in stan.

The first block in the stan file should be a function block that includes the implemented functions from WCPprior package.
```{stan,  output.var='model', eval=FALSE}
functions {
#include /include/WCP_analytic.stan
}
```

The data block specifies $N$ numbers of independent and identically distributed data with generalized Pareto distribution.
```{stan,  output.var='model', eval=FALSE}
data {
  int<lower=0> N; //number of data
  vector[N] y; //iid data
}

```

The parameter block specifies the two parameters as a two dimensional vector.
```{stan,  output.var='model', eval=FALSE}
parameters {
  real<lower = 0> sigma;
  real<lower = 0, upper = 1> xi;
}

```

The model block can be created by calling stan functions 'WCP1_2D_GP_analytic' and 'generalized_Pareto' from WCPprior package.
```{stan,  output.var='model', eval=FALSE}
model {
  target += WCP1_2D_GP_analytic_log(sigma, xi, 10);
  y ~ generalized_Pareto(sigma, xi);
}
```

Then the complete `stan` file is the following. It will be built and stored in the `model` object.

```{stan, output.var='model'}
functions {
#include ../inst/stan/include/WCP_analytic.stan
}
data {
  int<lower=0> N; //number of data
  vector[N] y; //iid data
}

parameters {
  real<lower = 0> sigma;
  real<lower = 0, upper = 1> xi;
}
model {
  target += WCP1_2D_GP_analytic_log(sigma, xi, 10);
  y ~ generalized_Pareto(sigma, xi);
}
```

Now we generate simulated generalized Pareto data. First we import a necessary library.
```{r,warning = FALSE, message=FALSE}
library("evmix")
```

We set $\sigma = 0.1, \xi = 2/3$ for the simulated data and call the function 'rgpd' to generate data.
```{r}
true_sigma = 0.1
true_xi = 2/3
sim_data = rgpd(n = 100, u = 0, sigmau = true_sigma, xi = true_xi, phiu = 1)
```


We will now fit the model in `stan`.
```{r, warning = FALSE, message=FALSE}
library(rstan)
fit_WCP_2D_GP <- sampling(model, data = list(N = 100, y = sim_data), iter = 1000) 
```


Let us plot the fitted model:

```{r}
plot(fit_WCP_2D_GP)
```
